### 2018-06-28 paper/2-background.tex
Work on S2

\begin{figure}[h]
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/make-example.pdf}}
\caption{Task dependency graph}
\end{subfigure}
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/make-example-full.pdf}}
\caption{Full rebuild}
\end{subfigure}
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/make-example-partial.pdf}}
\caption{Partial rebuild}
\end{subfigure}
\vspace{-2mm}
\caption{A task dependency graph and two build scenarios. Input files are shown
as rectangles, intermediate and output files are shown as rounded rectangles.
Modified inputs and files that are rebuilt are highlighted.
\label{fig-make}}
\vspace{-2mm}
\end{figure}

@@

To support dynamic dependencies, \Excel's build algorithm~\cite{excel_recalc} is
significantly different from \Make. \Excel arranges the cells into a linear
sequence, called the \emph{calc chain}.  During the build, \Excel processes
cells in the calc-chain sequence, but if computing a cell \cmd{C} requires the
value of a cell \cmd{D} that has not yet been computed, \Excel \emph{aborts}
computation of \cmd{C}, moves \cmd{D} before \cmd{C} in the calc chain, and
resumes the build starting with \cmd{D}. When a build is complete, the resulting
calc chain respects all the dynamic dependencies of the spreadsheet. When an
input value (or formula) is changed, \Excel uses the final calc chain from the
\emph{previous} build as its starting point so that, in the common case where
changing an input value does not change dependencies, there are no aborts.
Notice that build always succeeds regardless of the initial calc chain (barring
truly circular dependencies); the calc chain is just an optimisation.
We refer to this algorithm as \emph{restarting}, and will discuss it
in more detail in~\S\ref{sec-implementation-excel}.

@@

\Shake's implementation is different from both \Make and \Excel in two aspects.
First, it uses the dependency graph from the previous build to decide which
files need to be rebuilt. This idea has a long history, going back to
\emph{incremental}~\cite{demers1981incremental},
\emph{adaptive}~\cite{acar2002adaptive}, and
\emph{self-adjusting computations} (see \cite{acar2007selfadjusting} and
\S\ref{sec-related}). Second, instead of aborting and deferring the execution of
tasks whose newly discovered dependencies have not yet been built (as \Excel
does), \Shake \emph{suspends} their execution until the dependencies are brought
up to date. We refer to this build algorithm as \emph{suspending}.

@@

The user starts by checking out the project sources, whose hashes are (for
simplicity)~1,~2~and~3, and requests to build \cmd{main.exe}, see
Fig.~\ref{fig-bazel}(a,b). By looking up the global history of all previous
builds of the project\footnote{Here we ignore the issue of limited cloud storage
resources for the sake of simplicity; in practice, old entries are regularly
evicted from the storage, as further discussed in~\S\ref{sec-cloud-aspects}.},
the build system finds that someone has already compiled these exact sources
before and their resulting files \cmd{util.o} and \cmd{main.o} had hashes 4~and~5.
Similarly, the build system finds that the hash of the resulting \cmd{main.exe}
should be 6 and downloads the actual binary from the shared cloud storage, since
it is the end build product and must therefore be materialised.

In the second scenario, shown in Fig.~\ref{fig-bazel}(c), the user modifies the
source \cmd{util.c}, thereby changing its hash from~1~to~7. The build system
finds that nobody has ever compiled the new $\{\cmd{util.c}, \cmd{util.h}\}$
combination and must therefore build \cmd{util.o}, which results in changing its
hash from~4~to~8. The combination of hashes of \cmd{util.o} and
\cmd{main.o} has not been encountered before either, therefore the build system
first downloads \cmd{main.o} from the cloud and then builds \cmd{main.exe} by
linking the two object files. When the build is complete, the results
can be uploaded to the cloud for future reuse by other team members.

\Bazel is one of the first examples of openly available cloud build systems.
Like \Make, it does not support dynamic dependencies and can therefore benefit
from the simplicity of building tasks in a statically known topological order.
It is minimal and supports the early cutoff optimisation. To support cloud
builds, \Bazel maintains a \emph{content-addressable cache} that can be used to
fetch a previously built file given the hash of its content, and
\emph{dependency graphs from all previous builds}, annotated with observed file
hashes. The latter allows builds to bypass the execution of a task, by predicting the
hash of the result from the hashes of its dependencies, and subsequently
fetch the result from the cache. A concrete implementation is provided
in~\S\ref{sec-implementations}.

\begin{figure}[t]
\begin{subfigure}[b]{0.25\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
\caption{Checkout source files}
\end{subfigure}
\begin{subfigure}[b]{0.40\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-build.pdf}}
\caption{Build \cmd{main.exe}}
\end{subfigure}
\begin{subfigure}[b]{0.31\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-rebuild.pdf}}
\caption{Modify \cmd{util.c} and rebuild}
\end{subfigure}
\caption{A cloud build example: (a)~checkout sources, (b)~download \cmd{main.exe}
from the cloud and skip intermediate files (only their hashes are needed),
(c)~modify \cmd{util.c} and rebuild \cmd{main.exe}, which requires building
\cmd{util.o} (since nobody has compiled \cmd{util.c} before) and downloading
\cmd{main.o}. File hashes are shown inside circles.
\label{fig-bazel}}
\end{figure}

@@

We summarise differences between four discussed build systems in
Table~\ref{tab-summary}. The column \emph{`persistent build information'} refers
to the information that build systems persistently store between builds:
\begin{itemize}
    \item \Make stores file modification times, or rather, it relies on the file
    system to do that.
    \item \Excel stores one dirty bit per cell and the calculation chain from
    the previous build.
    \item \Shake stores the dependency graph discovered in the previous build,
    annotated with file content hashes for efficient checking of file changes.
    \item \Bazel stores all dependency graphs discovered in previous builds
    annotated with file hashes, and the content-addressable cache.
\end{itemize}

\begin{table}[h]
\smaller
\centering
\begin{tabular}{l||l|l||l|c|c|c}
\hline
$\!$Build system$\!$& Persistent build information   & Algorithm    & Dependencies & Minimal & Cutoff & Cloud$\!$\\\hline
$\!$\Make       $\!$& File modification times        & Topological  & Static       & Yes     & No     & No   $\!$\\
$\!$\Excel      $\!$& Dirty cells, calculation chain & Reordering   & Dynamic      & No      & No     & No   $\!$\\
$\!$\Shake      $\!$& Previous dependency graph      & Recursive    & Dynamic      & Yes     & Yes    & No   $\!$\\
$\!$\Bazel      $\!$& All dependency graphs, cache   & Topological  & Static       & Yes     & Yes    & Yes  $\!$\\\hline
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Summary of build system differences.\label{tab-summary}}
\vspace{-6mm}
\end{table}

In this paper we elucidate which build system properties are consequences of
specific implementation choices (metadata and algorithm), and how one can obtain
new build systems with desired properties by recombining parts of existing
implementations. As a compelling example, we demonstrate how to combine the
advantages of \Shake and \Bazel in a cloud build system with dynamic
dependencies, see~\S\ref{sec-implementation-cloud}.

>>>>

\begin{figure}[h]
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/make-example.pdf}}
\caption{Task dependency graph}
\end{subfigure}
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/make-example-full.pdf}}
\caption{Full rebuild}
\end{subfigure}
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/make-example-partial.pdf}}
\caption{Partial rebuild}
\end{subfigure}
\vspace{-2mm}
\caption{A task dependency graph and two build scenarios. Input files are shown
in rectangles, intermediate and output files are shown in rounded rectangles.
Modified inputs and files that are rebuilt are highlighted.
\label{fig-make}}
\vspace{-2mm}
\end{figure}

@@

To support dynamic dependencies, \Excel's build engine~\cite{excel_recalc} is
significantly different from \Make. \Excel arranges the cells into a linear
sequence, called the \emph{calc chain}.  During the build, \Excel processes
cells in the calc-chain sequence, but if computing a cell \cmd{C} requires the
value of a cell \cmd{D} that has not yet been computed, \Excel \emph{aborts}
computation of \cmd{C}, moves \cmd{D} before \cmd{C} in the calc chain, and
resumes the build starting with \cmd{D}. When a build is complete, the resulting
calc chain respects all the dynamic dependencies of the spreadsheet. When an
input value (or formula) is changed, \Excel uses the final calc chain from the
\emph{previous} build as its starting point so that, in the common case where
changing an input value does not change dependencies, there are no aborts.
Notice that build always succeeds regardless of the initial calc chain (barring
truly circular dependencies); the calc chain is just an optimisation.
We refer to this algorithm as \emph{restarting}, and will discuss it
in more detail in~\S\ref{sec-implementation-excel}.

@@

\Shake's implementation is different from both \Make and \Excel in two aspects.
First, it uses the dependency graph from the previous build to decide which
files need to be rebuilt. This idea has a long history, going back to
\emph{incremental}~\cite{demers1981incremental},
\emph{adaptive}~\cite{acar2002adaptive}, and
\emph{self-adjusting computations} (see \cite{acar2007selfadjusting} and
\S\ref{sec-related}). Second, instead of aborting and deferring the execution of
tasks whose newly discovered dependencies have not yet been built (as \Excel
does), \Shake \emph{suspends} their execution until the dependencies are brought
up to date. We refer to this task scheduling algorithm as \emph{suspending}.

@@

The user starts by downloading the project sources, whose content hashes are
(for simplicity)~1,~2 and~3, and requests to build \cmd{main.exe}, see
Fig.~\ref{fig-bazel}(a,b). By looking up the global history of all previous
builds of the project\footnote{Here we ignore the issue of limited cloud storage
resources for the sake of simplicity; in practice, old entries are regularly
evicted from the storage, as further discussed in~\S\ref{sec-cloud-aspects}.},
the build system finds that someone has already compiled these exact sources
before and the resulting files \cmd{util.o} and \cmd{main.o} had hashes
4~and~5, respectively. Similarly, the build system finds that the hash of the
resulting \cmd{main.exe} was 6 and downloads the actual binary from the shared
cloud storage -- it must be materialised, because it is the end build product.

In the second scenario, shown in Fig.~\ref{fig-bazel}(c), the user modifies the
source file \cmd{util.c}, thereby changing its hash from~1~to~7. The cloud
lookup of the new combination $\{\cmd{util.c}, \cmd{util.h}\}$ fails,
which means that nobody has ever compiled it. The build system must therefore
build \cmd{util.o}, which results in materialising it with a new hash~8. The
combination of hashes of \cmd{util.o} and \cmd{main.o} has not been encountered
before either, therefore the build system first downloads \cmd{main.o} from the
cloud and then builds \cmd{main.exe} by linking the two object files. When the
build is complete, the results can be uploaded to the cloud for future reuse by
other team members.

\Bazel is one of the first examples of openly available cloud build systems.
As of writing, it is not possible to express dynamic dependencies in
user-defined build rules, however some of the predefined build rules require
dynamic dependencies and the internal build engine can cope with them by using
a \emph{restarting} task scheduler, which is similar to that of \Excel but does
not use the calc chain. \Bazel is minimal and supports the early cutoff
optimisation.

To support cloud builds, \Bazel maintains (i) a \emph{content-addressable cache}
that can be used to download a previously built file given the hash of its
content, and (ii) the history of all executed build commands annotated with
observed file hashes. The latter allows the build engine to bypass the execution
of a task, by predicting the hash of the result from the hashes of its
dependencies, and subsequently download the result from the cache. A concrete
implementation is provided in~\S\ref{sec-implementations}.

\begin{figure}[t]
\begin{subfigure}[b]{0.25\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
\caption{Download source files}
\end{subfigure}
\begin{subfigure}[b]{0.40\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-build.pdf}}
\caption{Build \cmd{main.exe}}
\end{subfigure}
\begin{subfigure}[b]{0.31\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-rebuild.pdf}}
\caption{Modify \cmd{util.c} and rebuild}
\end{subfigure}
\caption{A cloud build example: (a)~download sources, (b)~build \cmd{main.exe}
by downloading it from the cloud and skipping intermediate files (only their
hashes are needed), (c)~modify \cmd{util.c} and rebuild \cmd{main.exe}, which
requires building \cmd{util.o} (nobody has compiled \cmd{util.c} before) and
downloading \cmd{main.o} (it is needed for linking \cmd{main.exe}). File hashes
are shown in circles, and non-materialised intermediates in dashed rounded
rectangles.
\label{fig-bazel}}
\end{figure}

@@

We summarise differences between four discussed build systems in
Table~\ref{tab-summary}. The column \emph{`persistent build information'} refers
to the information that build systems persistently store between builds:
\begin{itemize}
    \item \Make stores file modification times, or rather, it relies on the file
    system to do that.
    \item \Excel stores one dirty bit per cell and the calc chain from the
    previous build.
    \item \Shake stores the dependency graph discovered in the previous build,
    annotated with file content hashes for efficient checking of file changes.
    \item \Bazel stores the content-addressable cache and the history of all
    previous build commands annotated with file hashes. This information is
    shared among all users of the build system.
\end{itemize}

\begin{table}[h]
\smaller
\centering
\begin{tabular}{l||l|l||l|c|c|c}
\hline
$\!$Build system$\!$& Persistent build information & Algorithm   & Dependencies   & Minimal & Cutoff & Cloud$\!$\\\hline
$\!$\Make       $\!$& File modification times      & Topological & Static         & Yes     & No     & No   $\!$\\
$\!$\Excel      $\!$& Dirty cells, calc chain      & Restarting  & Dynamic        & No      & No     & No   $\!$\\
$\!$\Shake      $\!$& Previous dependency graph    & Suspending  & Dynamic        & Yes     & Yes    & No   $\!$\\
$\!$\Bazel      $\!$& Cloud cache, command history & Restarting  & Static$^{(*)}$ & Yes     & Yes    & Yes  $\!$\\\hline
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Summary of build system differences; $^{(*)}$predefined build rules can
have dynamic dependencies.\label{tab-summary}}
\vspace{-6mm}
\end{table}

In this paper we elucidate which build system properties are consequences of
specific implementation choices (stored metadata and task scheduling algorithm),
and how one can obtain new build systems with desired properties by recombining
parts of existing implementations. As a compelling example, we demonstrate how
to combine the advantages of \Shake and \Bazel in a cloud build system with
dynamic dependencies, see~\S\ref{sec-implementation-cloud}.
